### 2018-03-16 submission/manuscript.Rmd
add method changes --trying to make more clear

***Human and Mock Samples:***
A single fecal sample was obtained from 4 individuals who were part of the Enterics Research Investigational Network (ERIN). The processing and storage of these samples have been previously published [@erin_seekatz_2016]. Other than confirmation that none of these individuals had a *Clostridium difficle* infection, clinical data and other types of meta data were not utilized or accessed for this study. All samples were extracted using the MOBIO^TM^ PowerMag Microbiome RNA/DNA extraction kit (now Qiagen, MD, USA). The ZymoBIOMICS^TM^ Microbial Community DNA Standard (Zymo, CA, USA) was used in this study and is made up of *Pseudomonas aeruginosa*, *Escherichia coli*, *Salmonella enterica*, *Lactobacillus fermentum*, *Enterococcus faecalis*, *Staphylococcus aureus*, *Listeria monocytogenes*, and *Bacillus subtilis* at equal genomic DNA abundance (http://www.zymoresearch.com/microbiomics/microbial-standards/zymobiomics-microbial-community-standards).

***PCR Protocol:***
The five different HiFi DNA polymerases that were tested included AccuPrime^TM^ (ThermoFisher, MA, USA), KAPA HIFI (Roche, IN, USA), Phusion (ThermoFisher, MA, USA), Platinum (ThermoFisher, MA, USA), and Q5 (New England Biolabs, MA, USA). The PCR cycle conditions for Platinum and Accuprime followed a previously published protocol [@protocol_Kozich_2013] (https://github.com/SchlossLab/MiSeq_WetLab_SOP/blob/master/MiSeq_WetLab_SOP_v4.md). The HiFi DNA polymerase activation time was 2 minutes, unless a different activation was specified. For Kappa and Q5, the protocol previously published by Gohl and colleagues was used [@taq_Gohl_2016]. For Phusion, the company defined conditions were used except for extension time, where the Accuprime and Platinum settings were used.

Both fecal and mock samples cycle conditions started at 15 and increased by 5 up to 35 cycles with amplicons used at each 5-step increase for sequencing. The fecal PCR consisted of all 4 samples at 15, 20, 25, 30, and 35 cycles for each HiFi DNA polymerase (total samples = 100). The mock communities had 4 replicates at 15, 20, 25, and 35 cycles and 10 replicates for 30 cycles for all HiFi DNA polymerase (total samples = 130). No mock community sample had enough PCR product at 15 cycles for adequate 16S rRNA gene sequencing.

***Sequence Processing:***
The mothur software program was utilized for all sequence processing steps [@mothur_schloss_2009]. Generally, the protocol followed what has been previously published [@protocol_Kozich_2013] (https://www.mothur.org/wiki/MiSeq_SOP). Two major differences from the stated protocol were the use of VSEARCH instead of UCHIME for chimera detection and the use of the OptiClust algorithm instead of average neighbor for OTU generation at 97% similarity [@vsearch_Rognes_2016; @uchime_Edgar_2011; @opticlust_Westcott_2017]. Sequence error was determined using the 'seq.error' command on mock samples after chimera removal and classification to the RDP to remove non-bacterial sequences [@mothur_schloss_2009; @vsearch_Rognes_2016; @rdp_Cole_2013].

***Analysis Workflow:***
The total number of OTUs after sub-sampling was analyzed for both the fecal and mock community samples. Cycle dependent affects on Bray-Curtis indices were next assessed for the fecal samples looking at both overall cycle differences and within individual differences for the previous cycle (e.g. 20 versus 25, 25 versus 30, etc.). For these community based measures, the fecal samples were analyzed at 4 different sub-sampling sequence depths (1000, 5000, 10000, and 15000) while the mock community samples were analysed at 3 levels (1000, 5000, 10000). Next, based on these observations we wanted to analyze potential reasons for these differences. First, analysis of general sequence error rate, number of sequences with an error, and base substitution were assessed in the mock community for each HiFi DNA polymerase. After assessing these errors, the total number of chimeras was determined after sequence processing and correlations to the number of OTUs also assessed.

>>>>

***Human and Mock Samples:***
A single fecal sample was obtained from 4 individuals who were part of the Enterics Research Investigational Network (ERIN). The processing and storage of these samples were previously published [@erin_seekatz_2016]. Other than confirmation that none of these individuals had a *Clostridium difficle* infection, clinical data and other types of meta data were not utilized or accessed for this study. All samples were extracted using the MOBIO^TM^ PowerMag Microbiome RNA/DNA extraction kit (now Qiagen, MD, USA). The ZymoBIOMICS^TM^ Microbial Community DNA Standard (Zymo, CA, USA) was used for mock communities and was made up of *Pseudomonas aeruginosa*, *Escherichia coli*, *Salmonella enterica*, *Lactobacillus fermentum*, *Enterococcus faecalis*, *Staphylococcus aureus*, *Listeria monocytogenes*, and *Bacillus subtilis* at equal genomic DNA abundance (http://www.zymoresearch.com/microbiomics/microbial-standards/zymobiomics-microbial-community-standards).

***PCR Protocol:***
The five different HiFi DNA polymerases that were tested included AccuPrime^TM^ (ThermoFisher, MA, USA), KAPA HIFI (Roche, IN, USA), Phusion (ThermoFisher, MA, USA), Platinum (ThermoFisher, MA, USA), and Q5 (New England Biolabs, MA, USA). The PCR cycle conditions for Platinum and Accuprime followed a previously published protocol [@protocol_Kozich_2013] (https://github.com/SchlossLab/MiSeq_WetLab_SOP/blob/master/MiSeq_WetLab_SOP_v4.md). The HiFi DNA polymerase activation time was 2 minutes, unless a different activation was specified. For Kappa and Q5, a previously published protocol was used [@taq_Gohl_2016]. For Phusion, the company defined conditions were used except for extension time, where the Accuprime and Platinum settings were used.

The cycle conditions for both fecal and mock samples started at 15 and increased by 5 up to 35 cycles with amplicons used at each 5-step increase for sequencing. The PCR of fecal DNA samples consisted of all 4 samples at 15, 20, 25, 30, and 35 cycles for each HiFi DNA polymerase (total sample n=100). The mock communities had 4 replicates at 15, 20, 25, and 35 cycles and 10 replicates for 30 cycles for all HiFi DNA polymerases (total samples n=130). No mock community sample had enough PCR product at 15 cycles for adequate 16S rRNA gene sequencing.

***Sequence Processing:***
The mothur software program was used for all sequence processing steps [@mothur_schloss_2009]. The protocol has been previously published [@protocol_Kozich_2013] (https://www.mothur.org/wiki/MiSeq_SOP). Two major differences from the published protocol were the use of VSEARCH instead of UCHIME for chimera detection and the use of the OptiClust algorithm instead of average neighbor for OTU generation at 97% similarity [@vsearch_Rognes_2016; @uchime_Edgar_2011; @opticlust_Westcott_2017]. Sequence error was determined using the 'seq.error' command on mock samples before the 'pre.cluster' command, before chimera removal, and after chimera removal [@mothur_schloss_2009; @vsearch_Rognes_2016; @rdp_Cole_2013].

***Analysis Workflow:***
The total number of OTUs was analyzed after sub-sampling for both the fecal and mock community samples. For fecal samples, cycle dependent affects on Bray-Curtis indices were assessed for cycle group and within individual differences from the previous cycle (e.g. 20 versus 25, 25 versus 30, etc.). These community based measures for fecal samples were analyzed at 4 different sub-sampling sequence depths (1000, 5000, 10000, and 15000) while the mock community samples were analysed at 3 levels (1000, 5000, 10000). Based on these observations we analyzed potential reasons for these differences. Analysis of the mock community of each HiFi DNA polymerase for general sequence error rate, number of sequences with an error, base substitution, and numbers of chimeras were assessed before the 'pre.cluster' command, before chimera removal, and after chimera removal. Additionally, the correlation between the number of chimeras and the number of OTUs was also assessed before the 'pre.cluster' command, before chimera removal, and after chimera removal.

### 2018-03-16 submission/manuscript.tex
add method changes --trying to make more clear

\textbf{\emph{Human and Mock Samples:}} A single fecal sample was
obtained from 4 individuals who were part of the Enterics Research
Investigational Network (ERIN). The processing and storage of these
samples have been previously published (Seekatz et al., 2016). Other
than confirmation that none of these individuals had a \emph{Clostridium
difficle} infection, clinical data and other types of meta data were not
utilized or accessed for this study. All samples were extracted using
the MOBIO\textsuperscript{TM} PowerMag Microbiome RNA/DNA extraction kit
(now Qiagen, MD, USA). The ZymoBIOMICS\textsuperscript{TM} Microbial
Community DNA Standard (Zymo, CA, USA) was used in this study and is
made up of \emph{Pseudomonas aeruginosa}, \emph{Escherichia coli},
\emph{Salmonella enterica}, \emph{Lactobacillus fermentum},
\emph{Enterococcus faecalis}, \emph{Staphylococcus aureus},
\emph{Listeria monocytogenes}, and \emph{Bacillus subtilis} at equal
genomic DNA abundance
(\url{http://www.zymoresearch.com/microbiomics/microbial-standards/zymobiomics-microbial-community-standards}).

\textbf{\emph{PCR Protocol:}} The five different HiFi DNA polymerases
that were tested included AccuPrime\textsuperscript{TM} (ThermoFisher,
MA, USA), KAPA HIFI (Roche, IN, USA), Phusion (ThermoFisher, MA, USA),
Platinum (ThermoFisher, MA, USA), and Q5 (New England Biolabs, MA, USA).
The PCR cycle conditions for Platinum and Accuprime followed a
previously published protocol (Kozich et al., 2013)
(\url{https://github.com/SchlossLab/MiSeq_WetLab_SOP/blob/master/MiSeq_WetLab_SOP_v4.md}).
The HiFi DNA polymerase activation time was 2 minutes, unless a
different activation was specified. For Kappa and Q5, the protocol
previously published by Gohl and colleagues was used (Gohl et al.,
2016). For Phusion, the company defined conditions were used except for
extension time, where the Accuprime and Platinum settings were used.

Both fecal and mock samples cycle conditions started at 15 and increased
by 5 up to 35 cycles with amplicons used at each 5-step increase for
sequencing. The fecal PCR consisted of all 4 samples at 15, 20, 25, 30,
and 35 cycles for each HiFi DNA polymerase (total samples = 100). The
mock communities had 4 replicates at 15, 20, 25, and 35 cycles and 10
replicates for 30 cycles for all HiFi DNA polymerase (total samples =
130). No mock community sample had enough PCR product at 15 cycles for
adequate 16S rRNA gene sequencing.

\textbf{\emph{Sequence Processing:}} The mothur software program was
utilized for all sequence processing steps (Schloss et al., 2009).
Generally, the protocol followed what has been previously published
(Kozich et al., 2013) (\url{https://www.mothur.org/wiki/MiSeq_SOP}). Two
major differences from the stated protocol were the use of VSEARCH
instead of UCHIME for chimera detection and the use of the OptiClust
algorithm instead of average neighbor for OTU generation at 97\%
similarity (Edgar et al., 2011; Rognes et al., 2016; Westcott \&
Schloss, 2017). Sequence error was determined using the `seq.error'
command on mock samples after chimera removal and classification to the
RDP to remove non-bacterial sequences (Schloss et al., 2009; Cole et
al., 2013; Rognes et al., 2016).

\textbf{\emph{Analysis Workflow:}} The total number of OTUs after
sub-sampling was analyzed for both the fecal and mock community samples.
Cycle dependent affects on Bray-Curtis indices were next assessed for
the fecal samples looking at both overall cycle differences and within
individual differences for the previous cycle (e.g.~20 versus 25, 25
versus 30, etc.). For these community based measures, the fecal samples
were analyzed at 4 different sub-sampling sequence depths (1000, 5000,
10000, and 15000) while the mock community samples were analysed at 3
levels (1000, 5000, 10000). Next, based on these observations we wanted
to analyze potential reasons for these differences. First, analysis of
general sequence error rate, number of sequences with an error, and base
substitution were assessed in the mock community for each HiFi DNA
polymerase. After assessing these errors, the total number of chimeras
was determined after sequence processing and correlations to the number
of OTUs also assessed.

>>>>

\textbf{\emph{Human and Mock Samples:}} A single fecal sample was
obtained from 4 individuals who were part of the Enterics Research
Investigational Network (ERIN). The processing and storage of these
samples were previously published (Seekatz et al., 2016). Other than
confirmation that none of these individuals had a \emph{Clostridium
difficle} infection, clinical data and other types of meta data were not
utilized or accessed for this study. All samples were extracted using
the MOBIO\textsuperscript{TM} PowerMag Microbiome RNA/DNA extraction kit
(now Qiagen, MD, USA). The ZymoBIOMICS\textsuperscript{TM} Microbial
Community DNA Standard (Zymo, CA, USA) was used for mock communities and
was made up of \emph{Pseudomonas aeruginosa}, \emph{Escherichia coli},
\emph{Salmonella enterica}, \emph{Lactobacillus fermentum},
\emph{Enterococcus faecalis}, \emph{Staphylococcus aureus},
\emph{Listeria monocytogenes}, and \emph{Bacillus subtilis} at equal
genomic DNA abundance
(\url{http://www.zymoresearch.com/microbiomics/microbial-standards/zymobiomics-microbial-community-standards}).

\textbf{\emph{PCR Protocol:}} The five different HiFi DNA polymerases
that were tested included AccuPrime\textsuperscript{TM} (ThermoFisher,
MA, USA), KAPA HIFI (Roche, IN, USA), Phusion (ThermoFisher, MA, USA),
Platinum (ThermoFisher, MA, USA), and Q5 (New England Biolabs, MA, USA).
The PCR cycle conditions for Platinum and Accuprime followed a
previously published protocol (Kozich et al., 2013)
(\url{https://github.com/SchlossLab/MiSeq_WetLab_SOP/blob/master/MiSeq_WetLab_SOP_v4.md}).
The HiFi DNA polymerase activation time was 2 minutes, unless a
different activation was specified. For Kappa and Q5, a previously
published protocol was used (Gohl et al., 2016). For Phusion, the
company defined conditions were used except for extension time, where
the Accuprime and Platinum settings were used.

The cycle conditions for both fecal and mock samples started at 15 and
increased by 5 up to 35 cycles with amplicons used at each 5-step
increase for sequencing. The PCR of fecal DNA samples consisted of all 4
samples at 15, 20, 25, 30, and 35 cycles for each HiFi DNA polymerase
(total sample n=100). The mock communities had 4 replicates at 15, 20,
25, and 35 cycles and 10 replicates for 30 cycles for all HiFi DNA
polymerases (total samples n=130). No mock community sample had enough
PCR product at 15 cycles for adequate 16S rRNA gene sequencing.

\textbf{\emph{Sequence Processing:}} The mothur software program was
used for all sequence processing steps (Schloss et al., 2009). The
protocol has been previously published (Kozich et al., 2013)
(\url{https://www.mothur.org/wiki/MiSeq_SOP}). Two major differences
from the published protocol were the use of VSEARCH instead of UCHIME
for chimera detection and the use of the OptiClust algorithm instead of
average neighbor for OTU generation at 97\% similarity (Edgar et al.,
2011; Rognes et al., 2016; Westcott \& Schloss, 2017). Sequence error
was determined using the `seq.error' command on mock samples before the
`pre.cluster' command, before chimera removal, and after chimera removal
(Schloss et al., 2009; Cole et al., 2013; Rognes et al., 2016).

\textbf{\emph{Analysis Workflow:}} The total number of OTUs was analyzed
after sub-sampling for both the fecal and mock community samples. For
fecal samples, cycle dependent affects on Bray-Curtis indices were
assessed for cycle group and within individual differences from the
previous cycle (e.g.~20 versus 25, 25 versus 30, etc.). These community
based measures for fecal samples were analyzed at 4 different
sub-sampling sequence depths (1000, 5000, 10000, and 15000) while the
mock community samples were analysed at 3 levels (1000, 5000, 10000).
Based on these observations we analyzed potential reasons for these
differences. Analysis of the mock community of each HiFi DNA polymerase
for general sequence error rate, number of sequences with an error, base
substitution, and numbers of chimeras were assessed before the
`pre.cluster' command, before chimera removal, and after chimera
removal. Additionally, the correlation between the number of chimeras
and the number of OTUs was also assessed before the `pre.cluster'
command, before chimera removal, and after chimera removal.
